State-space analysis of a continuous gravitational wave source with a pulsar timing array: inclusion of the pulsar terms

### 2024-01-09 manuscript/manuscript_2/paper.tex
EoD

% Abstract of the paper
\begin{abstract}	
	 Pulsar timing arrays can detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries. The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked using a Kalman filter concurrently with the pulse frequency modulation induced by a gravitational wave from a single source. The modulation is the sum of terms proportional to the gravitational wave strain at the Earth and at every pulsar in the array. Here we generalize previous state-space formulations of the pulsar timing array problem to include the pulsar terms. The performance of the generalized Kalman filter is tested using astrophysically representative software injections in Gaussian measurement noise. It is shown that including the pulsar terms corrects for biases in the parameter estimates (especially the sky position of the source) and decreases the root-mean-square relative error in the estimated one-dimensional posteriors by to 99\%, for a loud gravitational wave source with characteristic wave strain $h_0 = 10^{-12}$. For a quiet gravitational wave signal with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in parameter accuracy, but the minimum detectable strain drops by $8\%$.
\end{abstract}

@@

\subsection{Reparametrisation of the pulsar terms} \label{sec:pulsar_term2}
In this paper, we generalize the Kalman filter analysis in K24 to include the pulsar terms. To do so, we define a new parameter 
\begin{equation}
	\chi^{(n)} = \left\{ \Omega \left[ 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \mod 2 \pi \ , \label{eq:chi_param}
\end{equation}
such that Equation \eqref{eq:g_func_trig} becomes
\begin{align}
	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
	& \times \Big \{\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
	&- \cos \left [-\Omega t +\Phi_0 + \chi^{(n)} \right ] \Big \} \ .
	\label{eq:g_func_trig_chi}
\end{align}
The reparametrisation in terms of $\chi^{(n)}$ treats the pulsar-dependent phase correction $\Omega \left[ 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}$ in the argument of the cosine of the pulsar term as a composite parameter to be inferred for each pulsar.

In principle, it is possible to disentangle $\Omega$, $\boldsymbol{n}$, and $d^{(n)}$

and infer them individually, if $N$ is large enough; for example, $\boldsymbol{n}$ appears independently in the phase of the pulsar term and in the denominator of the first line of Equation \eqref{eq:g_func_trig}. In practice, however, the likelihood surface is highly corrugated along the $d^{(n)}$-axis, and the prior on $d^{(n)}$ is broad compared to the wavelength $2\pi/\Omega$, so $d^{(n)}$ and hence $\Omega$ are not identifiable for reasonable $N$. By replacing $d^{(n)}$ with $\chi^{(n)}$, we obtain a smooth likelihood function, whose maximum is located efficiently and accurately by the nested sampler. This point is discussed in more detail in Appendix \ref{sec:psr_term_challenges}. The new parametrisation does not increase the dimension of the parameter space, because we trade $d^{(n)}$ for $\chi^{(n)}$. Once $\Omega$, ${\boldsymbol{n}}$, and $\chi^{(n)}$ are inferred, it is possible to solve for $d^{(n)}$, albeit not uniquely; $d^{(n)}$ can be inferred up to an integer multiple of the Doppler-shifted wavelength, due to the ${\rm mod} \, 2 \pi$ operation in Equation \eqref{eq:chi_param}. However, this ambiguity is unavoidable, whether we replace $d^{(n)}$ with $\chi^{(n)}$ or not, and occurs in every PTA analysis. The static parameters specific to the pulsars in the array (c.f. Equation \eqref{eq:psrparams}) are now
\begin{equation}
	\boldsymbol{\theta'}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\chi^{(n)} \right\}_{1\leq n \leq N} \ ,  \label{eq:psrparamsnew}
\end{equation}
whilst $\boldsymbol{\theta}_{\rm gw}$  remains unchanged. We define the complete set of $7 +5N$ static parameters to be inferred as $\boldsymbol{\theta'} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta'}_{\rm psr}$.  
%We exclude $\chi^{(n)}$ from membership of $\boldsymbol{\theta'}_{\rm psr}$ as the other elements of $\boldsymbol{\theta'}_{\rm psr}$ depend solely on the properties of the pulsar, with no dependence on the properties of the GW source. In contrast $\chi^{(n)}$ is a cross term that depends on both the individual pulsar and the GW. 
%

@@

\section{Parameter estimation} \label{sec:pe_and_ms} 
In this section, we apply the Kalman filter and nested sampler to calculate the joint posterior probability distribution $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$ and compare it to the known, injected values. The procedure is undertaken for multiple realisations of $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$, and we estimate $\boldsymbol{\theta'}$ independently for each realisation. The aims are (i) to demonstrate that the analysis scheme works, i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations; (ii) to give a preliminary sense of its accuracy, and (iii) to quantify the natural random dispersion in the one-dimensional posterior medians. Quantifying the dispersion is important since it is a practical measure of the scheme's accuracy when it is applied to real astronomical data, where the true parameter values and specific noise realisations are unknown. The injected $\boldsymbol{\theta}_{\rm gw}$ values are selected to be astrophysically representative, as per the top section of Table \ref{tab:parameters_and_priors}.The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$ are as specified in Section \ref{sec:rep_example1}. All injected static parameters $\boldsymbol{\theta}'$ are summarised in the second column of Table \ref{tab:parameters_and_priors}. The specification of the priors on $\boldsymbol{\theta'}$ is described in Appendix \ref{sec:set_priors} and summarised in the rightmost column of Table \ref{tab:parameters_and_priors}.

In Section \ref{sec:rep_smbh_source} we demonstrate how to generate synthetic data, namely noisy frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t)$, with which to validate the scheme. In Section \ref{sec:pe_and_ms} we apply the analysis scheme to the synthetic data to estimate $\boldsymbol{\theta'}$ for a  particular, arbitrary, representative choice of $\boldsymbol{\theta}_{\rm gw}$, i.e. the SMBHB system. The scheme is validated on multiple noise realisations of the pulsar process noise $\xi^{(n)}(t)$ and the detector measurement noise $\varepsilon^{(n)}(t)$ in order to test the scheme more strongly and quantify the  irreducible ``cosmic'' variance in the parameter estimates. In Section \ref{sec:parameter_space} we extend the tests of Section  \ref{sec:pe_and_ms}
to cover a broader parameter space and consider a range of astrophysically relevant SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$.

@@

 It is important to check how well the pulsar-term phase $\chi^{(n)}$ is estimated, since the reparameterization involving $\chi^{(n)}$ is a key feature of our approach.

 approach, as explained in Section \ref{sec:pulsar_term2}. In short, it turns out that $\chi^{(n)}$ is estimated accurately for all $n$.

 Figure \ref{fig:corner_plot_3} displays the results as a corner plot for the representative subset $\chi^{(1)}, \dots, \chi^{(5)}$ over the ten noise realisations, in the same style as Figure \ref{fig:corner_plot_1}. We do not display the corner plot for $\chi^{(6)}$, $\dots$, $\chi^{(47)}$ because it is too big. With the exception of the results for $\chi^{(2)}$ (discussed below), all histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values. There is no evidence for correlations between parameter pairs. Most phases are recovered unambiguously across all noise realisations; for the 47 $\chi^{(n)}$ parameters (i.e. not just the five plotted in Figure \ref{fig:corner_plot_3}), the injected value is  contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 383 out of the 470 possible combinations of the 47 parameters and ten realizations.

Rarely, $\chi^{(n)}$ is not estimated consistently across multiple realizations for some $n$, e.g $\chi^{(2)}$ (third row, third column of Figure \ref{fig:corner_plot_3}), corresponding to PSR J0340+4130. Out of all the pulsars in the array, this object matches most closely the direction to the synthetic GW source, with $\boldsymbol{n} \cdot \boldsymbol{q}^{(2)} = -0.96$,  $\cos \chi^{(2)} \approx 0.07$, and $g^{(2)}(t) \approx XYZ$ The inability to accurately infer $\chi^{(2)}$ improves with the signal-to-noise ratio (SNR); for $h_0 = 1 \times 10^{-12}$, $\chi^{(2)}$ is recovered accurately; see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}. \newline

\subsection{Timing parameters} \label{sec:timing_parameters}
Similar results are obtained for the 4$N$ parameters in $\boldsymbol{\theta}_{\rm psr}'$ besides $\chi^{(n)}$. The injected values are recovered unambiguously across the 10 noise realisations. For the sake of brevity we do not display the calculated posterior distributions, because inferring $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this paper and PTA analyses. In short, the estimates of $f_{\rm em}(t_1)$ and $\dot{f}_{\rm em}(t_1)$ are guided into narrow ranges by the narrow priors. [then add 1-2 sentences about sigma including giving some numbers on percentage accuracy (min and max)] We remind the reader that $\gamma^{(n)}$ is not estimated in this paper; typically it satisfies $\gamma^{(n)} \sim 10^{-5} T_{\rm obs}$ [refs from before], so its influence is muted. \newline

@@

\subsection{Exploring the SMBHB parameter domain} \label{sec:parameter_space} 
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/pp_plot_canonical500}
	\caption{Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e. PP plot). The injections are 200 simulated GW sources generated by drawing randomly five parameters in $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions in Table \ref{tab:parameters_and_priors}. Each coloured curve corresponds to a different parameter (see legend). The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR. The grey shaded contours label the $1\sigma$, $2\sigma$ and 3$\sigma$ confidence intervals. For parameters with well estimated posteriors, the PP curve should fall along the diagonal of unit slope. $\Omega$ and $\alpha$ are generally well-estimated (i.e. the curves lie close to the unit diagonal). The remaining three parameters, $\psi$, $\phi_0$ and $\delta$ show modest evidence of being over-constrained and stray outside the shaded region.}
	\label{fig:parameter_space}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/pp_plot_canonical_nochi}
	\caption{Same as Figure \ref{fig:parameter_space} but now assuming that $\chi^{(n)}$ are known exactly \textit{a priori} (i.e.with a delta-function prior for the nested sampling inference). The excursions from the shaded region seen in Figure  \ref{fig:parameter_space} are reduced; nearly all parameters, with the exception of $\delta$, are contained within the shaded region. }
	\label{fig:parameter_space2}
\end{figure}

@@

The PP results are displayed in Figure \ref{fig:parameter_space}. The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for 200 injections. The curves for the five parameters, colour-coded in the legend, are approximately linear. Some parameters are consistently well estimated across the parameter domain, e.g. the blue curve for $\Omega$ lies everywhere within the 2$\sigma$ shaded region. Conversely some parameters, e.g. $\psi$, stray outside the $1\sigma$ shaded region and show evidence of being slightly over-constrained; there are more injections contained within lower-value credible intervals than would be expected statistically, and fewer injections contained within higher-value credible intervals. This occurs due to the low SNR of the injected GW signal and the varying sensitivity of the specific PTA configuration used to different regions of the sky. \newline

Manually inspecting the individual corner plots for each of the 200 injections, we see that the nested sampler does not return a unimodal posterior in the rare event that the source is located unfavourably, with $\boldsymbol{n} \cdot \boldsymbol{q}^{(n)} \approx -1$, such that one cannot infer $\chi^{(n)}$ accurately (see Section \ref{sec:chi_estim}). All injected sources are ``observed" with the same settings such as $T_{\rm obs}$, $T_{\rm cad}$ and $n_{\rm live}$. In Figure \ref{fig:parameter_space2} we display a second PP plot, arranged identically to Figure \ref{fig:parameter_space}, but assuming that the injected values of $\chi^{(n)}$ are known exactly for the sake of testing, i.e. setting a delta-function prior on $\chi^{(n)}$. The excursions out of the shaded region diminish compared to Figure \ref{fig:parameter_space}, confirming the importance of accurate estimates of $\chi^{(n)}$.

@@

\section{Bias when neglecting the pulsar terms}\label{sec:earth_vs_psr}
In Section \ref{sec:rep_example1} we have validated the state-space analysis scheme when the pulsar terms are included in the inference model. In this section, critically, we compare the effect of including the pulsar-term in the model (i.e. using Equation \eqref{eq:g_func_trig_chi}), as opposed to the omitting the pulsar-term (i.e. Equation \eqref{eq:g_func_trig_earth}). We do this using the specific representative case of the individual quasi-monochromatic SMBHB source described in Table \ref{tab:parameters_and_priors}. We refer to the model where the pulsar-term is included alongside the Earth term as $\mathcal{M}_{\rm psr \& earth}$. We refer to the model where the pulsar-term is omitted as 
$\mathcal{M}_{\rm earth}$. Model $\mathcal{M}_{\rm psr \& earth}$ is parameterised by $\boldsymbol{\theta}'$. Model $\mathcal{M}_{\rm earth}$ is parameterised by $\boldsymbol{\theta}_{\rm earth} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}'_{\rm psr}$. Both models are applied to identical realisations of the data, generated using the procedure as described in Section  \ref{sec:synthetic_data}. The priors on the static parameters in each model are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for $\mathcal{M}_{\rm psr \& earth}$, as described in Appendix \ref{sec:set_priors}. \newline

@@

In this section apply we apply the Kalman filter in conjunction with nested sampling in order to infer the joint posterior distribution for the static parameters of each model. For $\mathcal{M}_{\rm earth}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_earth} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}_{\rm earth})$. The likelihood returned by the Kalman filter is then used by nested sampling in order to estimate $p(\boldsymbol{\theta}_{\rm earth} | \boldsymbol{Y})$. For $\mathcal{M}_{\rm psr \& earth}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_chi} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}')$. The likelihood returned by the Kalman filter is then used by nested sampling in order to estimate $p(\boldsymbol{\theta'} | \boldsymbol{Y})$. The settings of the nested sampler (for example the number of live points, see Appendix \ref{sec:nested_sampling}) are identical for both models. \newline

We consider two representative systems. The first is a ``low-SNR" system with $h_0 = 5 \times 10^{-15}$, i.e. the exact system described in Table \ref{tab:parameters_and_priors}. The second is a ``high-SNR" system which has the same static injection parameters, but with an increased $h_0 = 10^{-12}$. The ``high-SNR" system is considered in order to evaluate the parameter estimation accuracy, and in particular any potential bias, without confusion from the measurement noise. In order to enable a clear comparison between the posterior probability distributions calculated using the two models, in this section we consider a single noise realisation of the synthetic data $\boldsymbol{Y}$. The results quoted are consistent across different noise realisations. \newline

@@

\section{Choice of priors}\label{sec:set_priors}
To deploy the nested sampling outlined in Appendix \ref{sec:nested_sampling} and the workflow of Appendix \ref{sec:workflow} it is necessary to specify a Bayesian prior $\boldsymbol{\theta}$ on the static parameters. In this section we outline how the priors are chosen. The priors on each static parameter are summarised in Table \ref{tab:parameters_and_priors}.  \newline

>>>>

% Abstract of the paper
\begin{abstract}	
	 Pulsar timing arrays can detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries. The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked using a Kalman filter concurrently with the pulse frequency modulation induced by a gravitational wave from a single source. The modulation is the sum of terms proportional to the gravitational wave strain at the Earth and at every pulsar in the array. Here we generalize previous state-space formulations of the pulsar timing array problem to include the pulsar terms. The performance of the generalized Kalman filter is tested using astrophysically representative software injections in Gaussian measurement noise. It is shown that including the pulsar terms corrects for biases in the parameter estimates (especially the sky position of the source) and decreases the root-mean-square relative error in the estimated one-dimensional posteriors by to 99\%, for a loud gravitational wave source with characteristic wave strain $h_0 = 10^{-12}$. For a quiet gravitational wave signal with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in parameter accuracy, but the minimum detectable strain drops by $13\%$.
\end{abstract}

@@

\subsection{Reparametrisation of the pulsar terms} \label{sec:pulsar_term2}
In this paper, we generalize the Kalman filter analysis in K24 to include the pulsar terms. To do so, we define a new parameter 
\begin{equation}
	\chi^{(n)} = \left\{ \Omega \left[ 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \mod 2 \pi \ , \label{eq:chi_param}
\end{equation}
such that Equation \eqref{eq:g_func_trig} becomes
\begin{align}
	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
	& \times \Big \{\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
	&- \cos \left [-\Omega t +\Phi_0 + \chi^{(n)} \right ] \Big \} \ .
	\label{eq:g_func_trig_chi}
\end{align}
The reparametrisation in terms of $\chi^{(n)}$ treats the pulsar-dependent phase correction $\Omega \left[ 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}$ in the argument of the cosine of the pulsar term as a composite parameter to be inferred for each pulsar. In principle, it is possible to disentangle $\Omega$, $\boldsymbol{n}$, and $d^{(n)}$ and infer them individually, if $N$ is large enough; for example, $\boldsymbol{n}$ appears independently in the phase of the pulsar term and in the denominator of the first line of Equation \eqref{eq:g_func_trig}. In practice, however, the likelihood surface is highly corrugated along the $d^{(n)}$-axis, and the prior on $d^{(n)}$ is broad compared to the wavelength $2\pi/\Omega$, so $d^{(n)}$ and hence $\Omega$ are not identifiable for reasonable $N$. By replacing $d^{(n)}$ with $\chi^{(n)}$, we obtain a smooth likelihood function, whose maximum is located efficiently and accurately by the nested sampler. This point is discussed in more detail in Appendix \ref{sec:psr_term_challenges}. The new parametrisation does not increase the dimension of the parameter space, because we trade $d^{(n)}$ for $\chi^{(n)}$. Once $\Omega$, ${\boldsymbol{n}}$, and $\chi^{(n)}$ are inferred, it is possible to solve for $d^{(n)}$, albeit not uniquely; $d^{(n)}$ can be inferred up to an integer multiple of the Doppler-shifted wavelength, due to the ${\rm mod} \, 2 \pi$ operation in Equation \eqref{eq:chi_param}. However, this ambiguity is unavoidable, whether we replace $d^{(n)}$ with $\chi^{(n)}$ or not, and occurs in every PTA analysis. The static parameters specific to the pulsars in the array (c.f. Equation \eqref{eq:psrparams}) are now
\begin{equation}
	\boldsymbol{\theta'}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\chi^{(n)} \right\}_{1\leq n \leq N} \ ,  \label{eq:psrparamsnew}
\end{equation}
whilst $\boldsymbol{\theta}_{\rm gw}$  remains unchanged. We define the complete set of $7 +5N$ static parameters to be inferred as $\boldsymbol{\theta'} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta'}_{\rm psr}$.  
%We exclude $\chi^{(n)}$ from membership of $\boldsymbol{\theta'}_{\rm psr}$ as the other elements of $\boldsymbol{\theta'}_{\rm psr}$ depend solely on the properties of the pulsar, with no dependence on the properties of the GW source. In contrast $\chi^{(n)}$ is a cross term that depends on both the individual pulsar and the GW. 
%

@@

\section{Parameter estimation} \label{sec:pe_and_ms} 
In this section, we apply the Kalman filter and nested sampler to calculate the joint posterior probability distribution $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$ and compare it to the known, injected values. The procedure is undertaken for multiple realisations of $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$, and we estimate $\boldsymbol{\theta'}$ independently for each realisation. The aims are (i) to demonstrate that the analysis scheme works, i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations; (ii) to give a preliminary sense of its accuracy, and (iii) to quantify the natural random dispersion in the one-dimensional posterior medians. Quantifying the dispersion is important since it is a practical measure of the scheme's accuracy when it is applied to real astronomical data, where the true parameter values and specific noise realisations are unknown. The injected $\boldsymbol{\theta}_{\rm gw}$ values are selected to be astrophysically representative, as per the top section of Table \ref{tab:parameters_and_priors}.The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$ are as specified in Section \ref{sec:rep_example1}. All injected static parameters $\boldsymbol{\theta}'$ are summarised in the second column of Table \ref{tab:parameters_and_priors}. The specification of the priors on $\boldsymbol{\theta'}$ is described in Appendix \ref{sec:set_priors} and summarised in the rightmost column of Table \ref{tab:parameters_and_priors}. \newline

In Section \ref{sec:rep_smbh_source} we apply the analysis scheme to synthetic data and estimate $\boldsymbol{\theta}_{\rm gw}$ for a  particular, arbitrary, representative choice of $\boldsymbol{\theta}_{\rm gw}$, i.e. the SMBHB system. The scheme is validated on multiple noise realisations of the pulsar process noise $\xi^{(n)}(t)$ and the detector measurement noise $\varepsilon^{(n)}(t)$ in order to test the scheme more strongly and quantify the  irreducible ``cosmic'' variance in the parameter estimates. In Section \ref{sec:chi_estim} we inspect the estimates of  $\chi^{(n)}$. In Section \ref{sec:timing_parameters} we briefly discuss the estimates of the remaining 4N parameters in $\boldsymbol{\theta}_{\rm psr}$. In Section \ref{sec:parameter_space} we extend the tests across a broader parameter domain and consider a range of astrophysically relevant SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$.

@@

 It is important to check how well the pulsar-term phase $\chi^{(n)}$ is estimated, since the reparameterization involving $\chi^{(n)}$ is a key feature of our approach, as explained in Section \ref{sec:pulsar_term2}. In short, it turns out that $\chi^{(n)}$ is estimated accurately for all $n$. Figure \ref{fig:corner_plot_3} displays the results as a corner plot for the representative subset $\chi^{(1)}, \dots, \chi^{(5)}$ over the ten noise realisations, in the same style as Figure \ref{fig:corner_plot_1}. We do not display the corner plot for $\chi^{(6)}$, $\dots$, $\chi^{(47)}$ because it is too big. With the exception of the results for $\chi^{(2)}$ (discussed below), all histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values. There is no evidence for correlations between parameter pairs. Most phases are recovered unambiguously across all noise realisations; for the 47 $\chi^{(n)}$ parameters (i.e. not just the five plotted in Figure \ref{fig:corner_plot_3}), the injected value is  contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 383 out of the 470 possible combinations of the 47 parameters and ten realizations. \newline

Rarely, $\chi^{(n)}$ is not estimated consistently across multiple realizations for some $n$, e.g $\chi^{(2)}$ (third row, third column of Figure \ref{fig:corner_plot_3}), corresponding to PSR J0340+4130. Out of all the pulsars in the array, this object matches most closely the direction to the synthetic GW source, with $\boldsymbol{n} \cdot \boldsymbol{q}^{(2)} = -0.96$ and $\cos \chi^{(2)} \approx 0.07$. The inability to accurately infer $\chi^{(2)}$ improves with the signal-to-noise ratio (SNR); for $h_0 = 1 \times 10^{-12}$, $\chi^{(2)}$ is recovered unambiguously; see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}. \newline

\subsection{Timing parameters} \label{sec:timing_parameters}
Similar results are obtained for the 4$N$ parameters in $\boldsymbol{\theta}_{\rm psr}'$ besides $\chi^{(n)}$. The injected values are recovered unambiguously across the 10 noise realisations. For the sake of brevity we do not display the calculated posterior distributions, because inferring $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this paper and PTA analyses. In short, the estimates of $f_{\rm em}(t_1)$ and $\dot{f}_{\rm em}(t_1)$ are guided into narrow ranges by the narrow priors. The one-dimensional posteriors inferred for $\sigma^{(n)}$ are generally broader due to the broader prior, but contain the injection within the  90\% credible interval in a majority of cases. We remind the reader that $\gamma^{(n)}$ is not estimated in this paper; typically it satisfies $\gamma^{(n)} \sim 10^{-5} T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}, so its influence is muted. \newline

@@

\subsection{Exploring the SMBHB parameter domain} \label{sec:parameter_space} 
%\begin{figure}
%	\centering
%	\includegraphics[width=\columnwidth]{images/pp_plot_canonical500}
%	\caption{Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e. PP plot). The injections are 200 simulated GW sources generated by drawing randomly five parameters in $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions in Table \ref{tab:parameters_and_priors}. Each coloured curve corresponds to a different parameter (see legend). The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR. The grey shaded contours label the $1\sigma$, $2\sigma$ and 3$\sigma$ confidence intervals. For parameters with well estimated posteriors, the PP curve should fall along the diagonal of unit slope. $\Omega$ and $\alpha$ are generally well-estimated (i.e. the curves lie close to the unit diagonal). The remaining three parameters, $\psi$, $\phi_0$ and $\delta$ show modest evidence of being over-constrained and stray outside the shaded region.}
%	\label{fig:parameter_space}
%\end{figure}
%\begin{figure}
%	\centering
%	\includegraphics[width=\columnwidth]{images/pp_plot_canonical_nochi}
%	\caption{Same as Figure \ref{fig:parameter_space} but }
%	\label{fig:parameter_space2}
%\end{figure}

\begin{figure}
	\centering
	\subfloat[a][]{\includegraphics[width=\columnwidth]{images/pp_plot_canonical500} \label{fig:parameter_space}} \\
	\subfloat[b][]{\includegraphics[width=\columnwidth]{images/pp_plot_canonical_nochi} \label{fig:parameter_space2}}
	\caption{(a) Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e. PP plot). The injections are 200 simulated GW sources generated by drawing randomly five parameters in $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions in Table \ref{tab:parameters_and_priors}. Each coloured curve corresponds to a different parameter (see legend). The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR. The grey shaded contours label the $1\sigma$, $2\sigma$ and 3$\sigma$ confidence intervals. For parameters with well estimated posteriors, the PP curve should fall along the diagonal of unit slope. $\Omega$ and $\alpha$ are generally well-estimated (i.e. the curves lie close to the unit diagonal). The remaining three parameters, $\psi$, $\phi_0$ and $\delta$ show modest evidence of being over-constrained and stray outside the shaded region. (b) Same as (a) but now assuming that $\chi^{(n)}$ are known exactly \textit{a priori} (i.e.with a delta-function prior for the nested sampling inference). The excursions from the shaded region seen in Figure  \ref{fig:parameter_space} are reduced; nearly all parameters, with the exception of $\delta$, are contained within the shaded region. } \label{fig:AB}
\end{figure}

@@

The PP results are displayed in Figure \ref{fig:parameter_space}. The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for 200 injections. The curves for the five parameters, colour-coded in the legend, are approximately linear. Some parameters are consistently well estimated across the parameter domain, e.g. the blue curve for $\Omega$ lies everywhere within the 2$\sigma$ shaded region. Conversely some parameters, e.g. $\Phi_0$, stray outside the $1\sigma$ shaded region and show evidence of being slightly over-constrained; there are more injections contained within lower-value credible intervals than would be expected statistically, and fewer injections contained within higher-value credible intervals. This occurs due to the low SNR of the injected GW signal and the varying sensitivity of the specific PTA configuration used to different regions of the sky. More quantitatively, for the best estimated parameter $\Omega$ the injection is contained within the 90 \% credible interval in 91\% of cases. For the worst estimated parameters such as $\Phi_0$, the injection is contained within the 90 \% credible interval in 81\% of cases. \newline

Manually inspecting the individual corner plots for each of the 200 injections, we see that the nested sampler does not return a unimodal posterior in the rare event that the source is located unfavourably, with $\boldsymbol{n} \cdot \boldsymbol{q}^{(n)} \approx -1$, such that one cannot infer $\chi^{(n)}$ accurately (see Section \ref{sec:chi_estim}). All injected sources are ``observed" with the same settings such as $T_{\rm obs}$, $T_{\rm cad}$ and $n_{\rm live}$. In Figure \ref{fig:parameter_space2} we display a second PP plot, arranged identically to Figure \ref{fig:parameter_space}, but assuming that the injected values of $\chi^{(n)}$ are known exactly for the sake of testing, i.e. setting a delta-function prior on $\chi^{(n)}$. The excursions out of the shaded region diminish compared to Figure \ref{fig:parameter_space}, confirming the importance of accurate estimates of $\chi^{(n)}$. For the best estimated parameter $\Omega$ the injection is contained within the 90 \% credible interval in 91\% of cases, the same as in Figure \ref{fig:parameter_space}. For the worst estimated parameters such as $\Phi_0$, the injection is contained within the 90 \% credible interval in 89\% of cases, an improvement over the results of Figure Figure \ref{fig:parameter_space}. \newline

@@

\section{Bias when neglecting the pulsar terms}\label{sec:earth_vs_psr}
In Section \ref{sec:rep_example1} we have validated the state-space analysis scheme when the pulsar terms are included in the inference model. In this section, critically, we compare the effect of including the pulsar-term in the model (i.e. using Equation \eqref{eq:g_func_trig_chi}), as opposed to the omitting the pulsar-term (i.e. Equation \eqref{eq:g_func_trig_earth}). We do this using the specific representative case of the individual quasi-monochromatic SMBHB source described in Table \ref{tab:parameters_and_priors}. We refer to the model where the pulsar-term is included alongside the Earth term as $\mathcal{M}_{\rm psr \& earth}$. We refer to the model where the pulsar-term is omitted as 
$\mathcal{M}_{\rm earth}$. Model $\mathcal{M}_{\rm psr \& earth}$ is parameterised by $\boldsymbol{\theta}'$. Model $\mathcal{M}_{\rm earth}$ is parameterised by $\boldsymbol{\theta}_{\rm earth} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}'_{\rm psr}$. Both models are applied to identical realisations of the data, generated using the procedure as described in Section  \ref{sec:rep_example1}. The priors on the static parameters in each model are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for $\mathcal{M}_{\rm psr \& earth}$, as described in Appendix \ref{sec:set_priors}. \newline

@@

In this section we apply the Kalman filter in conjunction with nested sampling in order to infer the joint posterior distribution for the static parameters of each model. For $\mathcal{M}_{\rm earth}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_earth} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}_{\rm earth})$. The likelihood returned by the Kalman filter is then used by the nested sampler in order to estimate $p(\boldsymbol{\theta}_{\rm earth} | \boldsymbol{Y})$. For $\mathcal{M}_{\rm psr \& earth}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_chi} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}')$. The likelihood returned by the Kalman filter is then used by the nested sampler in order to estimate $p(\boldsymbol{\theta'} | \boldsymbol{Y})$. The settings of the nested sampler (for example the number of live points, see Appendix \ref{sec:nested_sampling}) are identical for both models. \newline

We consider two representative systems. The first is a ``low-SNR" system with $h_0 = 5 \times 10^{-15}$, i.e. the exact system described in Table \ref{tab:parameters_and_priors}. The second is a ``high-SNR" system which has the same static injection parameters, but with an increased strain amplitude $h_0 = 10^{-12}$. The ``high-SNR" system is considered in order to evaluate the parameter estimation accuracy, and in particular any potential bias, without confusion from the measurement noise. In order to enable a clear comparison between the posterior probability distributions calculated using the two models, in this section we consider a single noise realisation of the synthetic data $\boldsymbol{Y}$. The results quoted are consistent across different noise realisations. \newline

@@

\section{Choice of priors}\label{sec:set_priors}
To deploy the nested sampling outlined in Appendix \ref{sec:nested_sampling} and the workflow of Appendix \ref{sec:workflow} it is necessary to specify a Bayesian prior $\pi(\boldsymbol{\theta})$ on the static parameters. In this section we outline how the priors are chosen. The priors on each static parameter are summarised in Table \ref{tab:parameters_and_priors}.  \newline
